---
title: "Lab 2"
author: "Emil Luusua"
date: "25 September 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(HMM)
```


## Question 1

We create a Hidden Markov Model (HMM) with a hidden state Z_t that takes values z0-z9 which represent the sector where the robot is at time step t, and an observed random variable X_t that takes values x0-x9 which represent the reading of the tracking device at time step t. The transition probabilities are 0.5 to stay in the current state, and 0.5 to move to the next one. The emission probabilities are 0.2 to observe any of the states within 2 steps of the actual one.
```{r}
states <- c('z0', 'z1', 'z2', 'z3', 'z4', 'z5', 'z6', 'z7', 'z8', 'z9')
symbols <- c('x0', 'x1', 'x2', 'x3', 'x4', 'x5', 'x6', 'x7', 'x8', 'x9')
transProbs <- matrix(c(
  c(0.5, 0.5, 0, 0, 0, 0, 0, 0, 0, 0),
  c(0, 0.5, 0.5, 0, 0, 0, 0, 0, 0, 0),
  c(0, 0, 0.5, 0.5, 0, 0, 0, 0, 0, 0),
  c(0, 0, 0, 0.5, 0.5, 0, 0, 0, 0, 0),
  c(0, 0, 0, 0, 0.5, 0.5, 0, 0, 0, 0),
  c(0, 0, 0, 0, 0, 0.5, 0.5, 0, 0, 0),
  c(0, 0, 0, 0, 0, 0, 0.5, 0.5, 0, 0),
  c(0, 0, 0, 0, 0, 0, 0, 0.5, 0.5, 0),
  c(0, 0, 0, 0, 0, 0, 0, 0, 0.5, 0.5),
  c(0.5, 0, 0, 0, 0, 0, 0, 0, 0, 0.5)), 10)
emissionProbs <- matrix(c(
  c(0.2, 0.2, 0.2, 0, 0, 0, 0, 0, 0.2, 0.2),
  c(0.2, 0.2, 0.2, 0.2, 0, 0, 0, 0, 0, 0.2),
  c(0.2, 0.2, 0.2, 0.2, 0.2, 0, 0, 0, 0, 0),
  c(0, 0.2, 0.2, 0.2, 0.2, 0.2, 0, 0, 0, 0),
  c(0, 0, 0.2, 0.2, 0.2, 0.2, 0.2, 0, 0, 0),
  c(0, 0, 0, 0.2, 0.2, 0.2, 0.2, 0.2, 0, 0),
  c(0, 0, 0, 0, 0.2, 0.2, 0.2, 0.2, 0.2, 0),
  c(0, 0, 0, 0, 0, 0.2, 0.2, 0.2, 0.2, 0.2),
  c(0.2, 0, 0, 0, 0, 0, 0.2, 0.2, 0.2, 0.2),
  c(0.2, 0.2, 0, 0, 0, 0, 0, 0.2, 0.2, 0.2)), 10)

hmm <- initHMM(states, symbols, transProbs = transProbs, emissionProbs = emissionProbs)
```


## Question 2
```{r}
simHMM(hmm, 100)
```
